Variation in gaze understanding across the life span: A process-level perspective

### 2023-08-10 paper/manuscript.Rmd
adjust gaze model parts

keywords          : "social-cognitive development, theory of mind, gaze cues, individual differences, cognitive modeling, lifespan"
wordcount         : "xxx max 8000"
floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
header-includes   :
  - \usepackage{setspace}
  - \captionsetup[figure]{font={stretch=1}}
editor_options    : 
  markdown: 
    wrap: sentence
bibliography      : "../../references.bib"
bibliography: references.bib
---

@@

(ref:figlab1) **Differentiation in gaze understanding**. Developmental trajectory in gaze understanding across the lifespan. Performance is measured as imprecision, i.e., the absolute distance between the target's center and the participant's click (averaged across trials). The unit of imprecision is counted in the width of the target, i.e., a participant with imprecision of 1 clicked on average one target width to the left or right of the true target center. Grey dots show the mean performance for each subject averaged across trials. Blue lines show 100 draws from the expectation of the posterior predictive distribution (i.e, conditional expectation) of the Gaussian Process model, with its mean predicted developmental trajectory as a solid black line. Vertical, black, dashed lines show the location of the most prominent change points.

@@

The gaze model assumes participants use the agent's gaze as a cue to locate the target.[^3] Participants are assumed to observe the current location of the pupil and estimate the center of the agent's eyeball. Connecting these two point estimates as a line yields a gaze vector: the unique vector that extends from the center of the agent's eyeball through the center of the pupil to the attentional focal point. Participants are assumed to sample gaze vectors from a Normal distribution centered around the true gaze vector. They are assumed to do this independently for the left and right eye and then integrate the information to estimate the target's location. This inferential process involves some uncertainty about the true gaze vector.

@@

Now, individuals can differ in how wide their distribution around the true gaze vector is. The model's inferential component, α, estimates the standard deviation of the distribution for each individual participant. With higher standard deviations (i.e., increased α), participants are expected to show decreasing precision in estimating the target location. The model assumes that the sensitivity to the available gaze information changes with age.

The geometry of sampling from a distribution around the gaze vector leads to an interesting, testable group-level prediction. As the pupil location varies, a fixed amount of uncertainty around the gaze vector corresponds to a varying degree of uncertainty in the estimated target location. When the agent directs their gaze toward the very left or right side, the distribution around the gaze vector from which participants sample is comparatively wider than when the agent gazes centrally to the ground in front of them. For illustrative purposes, imagine a similar phenomenon: pointing a torch light to the ground. When one directs the torchlight straight onto the ground, the light beam is clearly defined and smaller than when one points the torchlight further away from oneself. Therefore, our gaze model predicts that our trials vary in difficulty: participants should be more imprecise in locating the target the further out it lands.[^4] If our data matches the pattern of this model prediction, this can act as evidence for the gaze model. Therefore, our gaze model provides a quantitative theory of gaze understanding, with testable model predictions.

[^4]: In our screen-based study, this effect should decrease again towards the most outward sides. Since the computer screen has a natural border, trials in which the target lands furthest out to the left/right become slightly easier again. In these cases, the uncertainty about the true gaze vector is - practically - only on the inner side (facing the center) of the screen. The natural border of the screen limits where participants can respond. In another adult sample with more trials, we could recover this pattern. For further elaboration, see Supplements.

@@

Our goal was to describe the inferential process of gaze understanding. We quantified how well our model explained the gaze understanding process by comparing it to two alternative models that make different assumptions of which information children use and where they consequently click to locate the target. Our modeling framework consisted of three mutually exclusive models: (1) a gaze model, (2) a random guessing model, and (3) a center bias model. We gauged which model can best explain our data by conducting model comparisons. All cognitive models and model comparisons were implemented in WebPPL [@goodman2014design].

To gauge the plausibility of our gaze model, we implemented two models that represent alternatives about how participants solve the TANGO. Participants who were overall very imprecise in locating the target might be less likely to use the agent's gaze as a cue at all. The alternative models, therefore, do not assume that participants made use of the gaze cue. The first alternative model assumed participants were randomly guessing. This was implemented as sampling from a uniform distribution over all possible coordinates (i.e., 0 - 1920). The second alternative model assumed the participants were biased towards the screen center: Participants could be drawn toward the screen center since the agent and the starting point of the balloon were located there. This bias was implemented as sampling from a Normal distribution, with the center of the screen as the mean and one target width as the standard deviation.

@@

@@

For the gaze model, we found a high similarity between the model predictions and the observed data across the entire age range. The developmental trajectory that we observed in the data closely resembled the predictions of the model: the correlation between the data and the model predictions was XXX [TODO]. The model explained XXX% [TODO] of the variance in the data.

@@

(ref:figlab2) **Gaze model**

@@

```{r magnet_results}
magnet_williamsttest_tom <- readRDS("../saves/williams_tom.rds")
magnet_williamsttest_perspective <- readRDS("../saves/williams_perspective.rds")
magnet_cormatrix <- readRDS("../saves/magnet_cormatrix.rds")
magnet_model <- readRDS("../saves/magnet_winningmodel.rds")
```

@@

(ref:figlab3) **Components of gaze understanding.** (A) Study procedures. Top: TANGO (i.e., social vector estimation; gaze understanding). Bottom: Magnet (i.e., non-social vector estimation). Left: Screenshots of familiarization phase. Right: Screenshots of test phase. (B) Magnet task: non-social vector estimation.(B) Correlations between gaze understanding, physical vector estimation, ToM, and perspective-taking. Dots show the correlation coefficients, while error bars represent 95% CIs. (C) Influence of perspective-taking and physical vector estimation on gaze understanding. The graph show the posterior distributions for the respective predictor. Black dots represent means, thicker black lines 80% CrI and thinner black lines 95% CrI.

>>>>

keywords          : "social-cognitive development, theory of mind, gaze cues, individual differences, cognitive modeling, lifespan"
wordcount         : "xxx max 8000"
floatsintext      : yes
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : yes
mask              : yes 
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_word
header-includes   :
  - \usepackage{setspace}
  - \captionsetup[figure]{font={stretch=1}}
editor_options    : 
  markdown: 
    wrap: sentence
bibliography      : "../../references.bib"
---

@@

(ref:figlab1) **Developmental trajectory of gaze understanding across the lifespan**. Performance is measured as imprecision, i.e., the absolute distance between the target's center and the participant's click (averaged across trials). The unit of imprecision is counted in the width of the target, i.e., a participant with imprecision of 1 clicked on average one target width to the left or right of the true target center. Grey dots show the mean performance for each subject averaged across trials. Blue lines show 100 draws from the expectation of the posterior predictive distribution (i.e, conditional expectation) of the Gaussian Process model, with its mean predicted developmental trajectory as a solid black line. Vertical, black, dashed lines show the location of the most prominent change points.

@@

Our model quantifies a participant's cognitive ability to follow gaze by inverting a probabilistic process that generates the participant's clicks from observing the eyes of the agent. It is formally defined as:

\begin{equation}
    P(\mathbf{\theta} | x_c, \alpha_l, \alpha_r) \propto P(x_c | \alpha_l, \alpha_r, \mathbf{\theta})P(\mathbf{\theta})
\end{equation}

where $\mathbf{\theta}$ is an individual's cognitive ability to locate the focus of the agent's attention, $x_c$ is the coordinate the participant clicked, and $\alpha_l$ and $\alpha_r$ are the pupil angles for the left and right eye, respectively. The pupil angle is defined as the angle between a line connecting the center of the eye to the pupil and a line extended vertically downward from the pupil.[^3]

@@

Based on our verbal task instructions, we assume that participants (1) expect the agent's looks to be directed at the target, and (2) to click on the coordinate they estimate the agent to look at. 
Consequently, we do not assume that participant's clicks are noisy in any way but that they click on the screen location where they genuinely think the target is (and that the agent is looking at). 
However, the true eye angles ($\alpha_l$ and $\alpha_r$) cannot be directly observed. These have to be estimated based on the position of the pupils within the eyes, resulting in approximate values ($\hat{\alpha_l}$ and $\hat{\alpha_r}$).
We presume this estimation to be a noisy process. Thus, the development of the cognitive ability to follow gaze corresponds to a reduction in the magnitude of the noise in the estimates (i.e., an increased certainty about the eye angles).

Any clicked value of $x_c$ implies a "matched pair" of the estimated pupil angles $\hat{\alpha_l}$ and $\hat{\alpha_r}$, with the property that lines extended along those two angles meet at the precise location of the target. 
As a consequence, we can rewrite the likelihood function of the model above:

\begin{equation}
P(x_c | \alpha_l, \alpha_r, \theta) \propto P(\hat{\alpha_l}, \hat{\alpha_r} | \alpha_l, \alpha_r, \theta) P(x_c)
\end{equation}

The second term of the right-hand side equation above, $P(x_c)$, is a prior over potential target locations, which we assume to be skewed towards the screen center: 
We anticipate that participants have an a priori expectation that the target will land close to the middle, partly because the target was last visible in the screen center before disappearing behind the hedge, and because the agent is located centrally on the screen. 
We estimate the strength of this center bias (i.e., the standard deviation of a Normal distribution around the center of the screen) based on the data.

The main inferential task for the participant lies in estimating the pupil angles, i.e., sampling from the first term of the right-hand side equation above, $P(\hat{\alpha_l}, \hat{\alpha_r} | \alpha_l, \alpha_r, \theta)$.
For this, we assume that the pair of estimated angles are sampled from a probability distribution which is the product of two Normal distributions of equal variance $\sigma_v$ centered on the true pupil angles:

\begin{equation}
P(\hat{\alpha_l}, \hat{\alpha_r} | \alpha_l, \alpha_r, \theta) \propto \phi(\hat{\alpha}_l ; \alpha_l, \sigma_v)\phi(\hat{\alpha}_r ; \alpha_r, \sigma_v),
\end{equation}

Broadly summarizing: Participants are assumed to observe the pupil locations and estimate the center of the agent's eyes.
Connecting these two point estimates as a line yields the unique vector that extends from the center of the agent's eyeball through the center of the pupil to the attentional focal point. 
Taking the angle between this vector and a line pointing vertically to the ground yields the pupil angle. 
Participants are assumed to sample from Normal distributions centered around the true pupil angle.
They are assumed to do this independently for the left and right eye and then integrate the information to estimate the target's location.

As $\sigma_v$ determines the level of accuracy with which the participant estimates the pupil angles, it is the component that defines $\mathbf{\theta}$.
When $\sigma_v$ is very small (i.e., the distribution around the pupil angle is narrow), clicks far away from the target are unlikely, as these would require estimated pupil angles very different from the true pupil angles.  
When $\sigma_v$ is very large (i.e., the distribution around the pupil angle is wide), almost any pupil angles may be sampled, corresponding to a roughly uniform distribution over click coordinates.
We expect $\sigma_v$ to vary between individuals. Consequently, individuals differ in the level of precision with which they can locate the target based on observing the agent's eyes.

The shape of the $P(x_c | \alpha_l, \alpha_r, \theta)$ distribution leads to an interesting, testable group-level prediction. 
As the pupil location varies, a fixed amount of uncertainty around the pupil angle corresponds to a varying degree of uncertainty in the estimated target location.
When the agent directs their gaze toward the very left or right side, the distribution around the pupil angle from which participants sample is comparatively wider than when the agent gazes centrally to the ground in front of them.
For illustrative purposes, imagine a similar phenomenon: pointing a torch light to a flat surface on the ground.
When one points the light cone directly at the surface, the light beam is concentrated in a clearly defined, small, symmetric area. 
When one points the light cone further away from oneself (shining at an angle), the light from one half of the cone must travel further to reach the surface than the light from the other half, resulting in an asymmetric light pattern. 
As the angle increases, the light is spread over a wider area, and the surface is illuminated less evenly. 
Consequently, for the same $\sigma_v$, the further out a target coordinate lies, the wider and less symmetric the distribution. This increases both the variance and the bias in a participant's estimate of the agent's attentional focus, resulting in decreased performance in the task. 
As $\sigma_v$ decreases and the cone narrows, the extent to which performance varies at different angles decreases.
Therefore, our gaze model predicts that our trials vary in difficulty: participants should be more imprecise in locating the target the further out it lands.[^4]
If our data matches the pattern of this model prediction, this can act as evidence for the gaze model.
Therefore, our gaze model provides a quantitative theory of gaze understanding, with testable model predictions.

[^4]: In our screen-based study, this effect should decrease again towards the most outward sides.
    Since the computer screen has a natural border, trials in which the target lands furthest out to the left/right become slightly easier again.
    In these cases, the uncertainty about the true gaze vector faces - practically - only the inner side (facing the center) of the screen, since the natural border of the screen limits where participants can click.
    In another adult sample with more trials, we could recover this pattern.
    For further elaboration, see Supplements.

@@

Our goal was to describe the inferential process of gaze understanding.
We quantified how well our gaze model explained the gaze understanding process by comparing it to two alternative models that make different assumptions of which information participants use and where they consequently click to locate the target.
Our modeling framework consisted of three mutually exclusive models: (1) a gaze model, (2) a random guessing model, and (3) a center bias model.
We gauged which model can best explain our data by conducting model comparisons.
All cognitive models were implemented in WebPPL [@goodman2014design].

To gauge the plausibility of our gaze model, we implemented two models that represent alternatives about how participants solve the TANGO.
Participants who were overall very imprecise in locating the target might be less likely to use the agent's gaze as a cue at all.
The alternative models, therefore, do not assume that participants made use of the gaze cue.
The first alternative model assumed participants were randomly guessing.
This was implemented as sampling from a uniform distribution over all possible coordinates (i.e., 0 - 1920).
The second alternative model assumed the participants always want to click at the screen center: Participants could be drawn toward the screen center since the agent and the starting point of the balloon were located there.
This was implemented as sampling from a Normal distribution, with the center of the screen as the mean and one target width as the standard deviation.

@@

## Results
```{r gazemodel_corrs}
gazemodel_corrs <- readRDS("../saves/gazemodel_correlations.rds")
```

@@

When correlating the observed data across all target positions with the predictions of the three models, we found a high similarity for the gaze model: *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_pred"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_pred"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_pred"]`], while the correlations with the alternative models were substantially smaller (center bias model: *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "centermodel_pred"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "centermodel_pred"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "centermodel_pred"]`]; random guessing model: *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "randommodel_pred"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "randommodel_pred"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "randommodel_pred"]`]).
The developmental trajectory of the estimated gaze model parameter also correlated highly with the observed data: *r* = `r gazemodel_corrs$r[gazemodel_corrs$corr == "gazemodel_param"]`, 95%CI [`r gazemodel_corrs$lower[gazemodel_corrs$corr == "gazemodel_param"]`, `r gazemodel_corrs$upper[gazemodel_corrs$corr == "gazemodel_param"]`].

@@

(ref:figlab2) **Gaze model** (A) Visualization of the gaze model. \alpha is defined as the pupil angle, i.e., the angle between a line connecting the eye center to the pupil and a line extended vertically downward from the pupil. $\sigma_v$ is the variance around the Normal distribution centered on the true pupil angle. This component is expected to vary between participants. (B) Developmental trajectory of the estimated model parameter. Grey dots show individual level parameter values. The black line shows the maximum a posteriori (MAP) estimate; blue lines show 1000 draws from it. (C) Correlation between estimated mode of the model parameter and data mean per individual, color-coded by age. The grey regression line with 95% CI shows smooth conditional mean based on a linear model, with *Pearson*'s correlation coefficient *r*. (D) Geometrical features of the gaze model. As the pupil location varies, a fixed amount of uncertainty around the gaze vector corresponds to a varying degree of uncertainty in the estimated target location. Top: Agent gazes centrally to the ground. Bottom: Agent gazes toward the side. The distribution around the gaze vector from which participants sample is wider compared to when the agent gazes centrally. The blue line on the ground shows the added level of uncertainty in the estimated target position for the target location further outward. (E) Pattern recovery. Imprecision in target width for each target bin by age group. Model predictions in blue; data in grey. (F) Correlation between the observed data and the predictions of the three models by target position (across age and individuals).

@@

```{r magnet_results}
magnet_williamsttest_tom <- readRDS("../saves/magnet_williams_tom.rds")
magnet_williamsttest_perspective <- readRDS("../saves/magnet_williams_perspective.rds")
magnet_cormatrix <- readRDS("../saves/magnet_cormatrix.rds")
magnet_model <- readRDS("../saves/magnet_winningmodel.rds")
```

@@

(ref:figlab3) **Components of gaze understanding.** (A) Study procedures. Top: TANGO (i.e., gaze understanding; social vector estimation). Bottom: Magnet (i.e., non-social vector estimation). Left hand side show screenshots of familiarization phase; right hand side show screenshots of the test phase. (B) Correlations between gaze understanding, physical vector estimation, ToM, and perspective-taking. Dots show the correlation coefficients, while error bars represent 95% CIs. (C) Influence of age, perspective-taking and physical vector estimation on gaze understanding. The graph shows the posterior distributions for the respective predictor. Black dots represent means, thicker black lines 80% CrI and thinner black lines 95% CrI.
